---
title: "project4"
author: "Brendon Happ"
date: "December 30, 2017"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(knitr)
library(gridExtra)
stroopdata <- read.csv("stroopdata.csv")
```

# Statistics: The Science of Decisions 

## 1. What is our independent variable? What is our dependent variable?
The independent variable is controlled so it would be the congruent word 
condition.

The dependent changes with changes to the independent variable, so that would be
the time.

## 2. What is an appropriate set of hypotheses for this task? What kind of statistical test do you expect to perform? Justify your choices. 

The goal is to see if it takes more time on average to read incongruent words or 
not. This can be described with the following hypotheses:

Null Hypothesis ($H_o$):
There is no statistically significant difference in time from reading congruent
and incongruent words.

$H_o$ : $\mu_{incongruent}$ = $\mu_{congruent}$

Alternate Hypothesis ($H_A$):
There is a statistically significant difference in time from reading congruent
and incongruent words.

$H_A$ : $\mu_{incongruent}$ > $\mu_{congruent}$

  Where:
  
    $\mu_{incongruent}$ = Average of time taken to read incongruent words
    $\mu_{congruent}$ = Average of time taken to read congruent words

From the data, we do not have the population standard deviation $\sigma$ and 
there are under 30 datasets. We also know that both of the tests were given to
the same people. Since the goal is to see if it took more time or not to read 
incongurent words, a one tailed dependent sample t test will be used.

## 3. Report some descriptive statistics regarding this dataset. Include at least one measure of central tendency and at least one measure of variability.

```{r, message=FALSE, warning=FALSE}
congruent.mean <- mean(stroopdata$Congruent) #14.051125
incongruent.mean <- mean(stroopdata$Incongruent) #22.015917
congruent.sd <- sd(stroopdata$Congruent) #3.559
incongruent.sd <- sd(stroopdata$Incongruent) #4.797
```
The average time to complete the task for congruent words was 14.05 seconds. The
average time for incongruent words is 22.016 seconds. The incongruent words 
varied more with a standard deviation of 4.797 versus 3.559 for congruent words.

## 4. Provide one or two visualizations that show the distribution of the sample data. Write one or two sentences noting what you observe about the plot or plots.

```{r, message=FALSE, warning=FALSE}
plot1 <- ggplot(stroopdata, aes(Congruent)) +geom_histogram(fill = 'red', 
                                                   color = 'white',
                                                   binwidth = 2.5)

plot2 <- ggplot(stroopdata, aes(Incongruent)) + geom_histogram(fill = 'green', 
                                                   color = 'white',
                                                   binwidth = 3.5)
grid.arrange(plot1, plot2)
```

The histograms above show that the data is normally distributed. They also show 
that the Incongruent words generally take a longer time than the Congruent words

## 5. Now, perform the statistical test and report your results. What is your confidence level and your critical statistic value? Do you reject the null hypothesis or fail to reject it? Come to a conclusion in terms of the experiment task. Did the results match up with your expectations?

```{r, message=FALSE, warning=FALSE}
n <- dim(stroopdata)[1] #24
alpha = .05
mean.diff <- incongruent.mean - congruent.mean #7.96479
SE <- sd(stroopdata$Incongruent - stroopdata$Congruent)/sqrt(n) #.993029
t <- mean.diff/SE #8.0207
t.crit <- qt(1-alpha, df = n-1) #1.71387
p <- pt(-abs(t), df = n-1) #2.5e^-8
con.low = mean.diff-(t.crit*SE) #6.26
con.high = mean.diff+(t.crit*SE) #9.67
```

These statistical results tell us that we will reject the null. The t critical 
value (1.71387) is smaller than the t statistic value (8.0207) at an $\alpha$ of 
.05. The p value (2.5e6-8) is also much smaller than $\alpha$, therefor we 
reject the null. This means we accept the alternative hypothesis and
$H_A$ : $\mu_{incongruent}$ > $\mu_{congruent}$, is true. The time it takes 
on average to read incongruent words is more than it would take for congruent
words.
The confidence interval on the mean difference is: 95% CI = (6.26, 9.67)

